---
title: "Lab 5"
author: "Erin Cagle"
format: html
editor: visual
embed-resources: true
---

# Learning goals

-   Use the `merge()` function to join two datasets.
-   Deal with missings and impute data.
-   Identify relevant observations using `quantile()`.
-   Practice your GitHub skills.

# Lab description

For this lab we will be, again, dealing with the meteorological dataset downloaded from the NOAA, the `met`. In this case, we will use `data.table` to answer some questions regarding the `met` dataset, while at the same time practice your Git+GitHub skills for this project.

This markdown document should be rendered using `github_document` document.

# Part 1: Setup the Git project and the GitHub repository

1.  Go to your documents (or wherever you are planning to store the data) in your computer, and create a folder for this project, for example, "PM566-labs"

2.  In that folder, save [this template](https://raw.githubusercontent.com/USCbiostats/PM566/master/website/content/assignment/05-lab.Rmd) as "README.Rmd". This will be the markdown file where all the magic will happen.

3.  Go to your GitHub account and create a new repository, hopefully of the same name that this folder has, i.e., "PM566-labs".

4.  Initialize the Git project, add the "README.Rmd" file, and make your first commit.

5.  Add the repo you just created on GitHub.com to the list of remotes, and push your commit to origin while setting the upstream.

Once you are done setting up the project, you can now start working with the MET data.

## Setup in R

1.  Load the `data.table` (and the `dtplyr` and `dplyr` packages if you plan to work with those).

    ```{r}
    library(data.table)
    library(dtplyr)
    library(dplyr)
    library(ggplot2)
    library(tibble)
    library(tidyr)
    library(purrr)
    library(forcats)
    ```
2.  Load the met data from https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz, and also the station data. For the later, you can use the code we used during lecture to pre-process the stations data:

    ```{r}
    if (!file.exists("met_all.gz"))
      download.file(
        url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
        destfile = "met_all.gz",
        method   = "libcurl",
        timeout  = 60
        )
    met <- data.table::fread("met_all.gz")
    ```
```{r stations-data, eval = FALSE}
# Download the data
stations <- fread("ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
stations[, USAF := as.integer(USAF)]

# Dealing with NAs and 999999
stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

# Selecting the three relevant columns, and keeping unique records
stations <- unique(stations[, list(USAF, CTRY, STATE)])

# Dropping NAs
stations <- stations[!is.na(USAF)]

# Removing duplicates
stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]

#3. Merge the data as we did during the lecture.
stat <- merge(
  # Data
  x     = met,      
  y     = stations, 
  # List of variables to match
  by.x  = "USAFID",
  by.y  = "USAF", 
  # Which obs to keep?
  all.x = TRUE,      
  all.y = FALSE
  )
head(stat[, list(USAFID, WBAN, STATE)], n = 4)
```
## Question 1: Representative station for the US

What is the median station in terms of temperature, wind speed, and atmospheric pressure? Look for the three weather stations that best represent continental US using the `quantile()` function. Do these three coincide?

```{r}
stat_med <- stat[, .(
    temp_quan      = quantile(temp, na.rm=TRUE),
    wind.sp_quan   = quantile(wind.sp, na.rm=TRUE),
    atm.press_quan = quantile(atm.press, na.rm = TRUE)
    ), by = USAFID
    ][order(USAFID)]
    

# in_names  <- c("wind.sp", "temp", "atm.press")
# out_names <- paste0(in_names, "_med")
# stat_med <- stat[,
#   setNames(lapply(.SD, quantile, na.rm = TRUE), out_names),
#   .SDcols = in_names
#   ]

usmed <- stat_med[, .(
    temp_quan      = quantile(temp_quan, na.rm=TRUE),
    wind.sp_quan   = quantile(wind.sp_quan, na.rm=TRUE),
    atm.press_quan = quantile(atm.press_quan, na.rm = TRUE))]
medstate <- stat[(temp == 23.900 & wind.sp == 2.6 & atm.press == 1014.200)]
medstate[order(medstate$STATE)]
```
Median temp is 23.9, median wind speed is 2.6, and median atm pressure is 1014.2. States that align with the median values are most often states in the southeast region.

```{r}
stat_med_state <- stat[, .(
    temp_quan      = quantile(temp, na.rm=TRUE),
    wind.sp_quan   = quantile(wind.sp, na.rm=TRUE),
    atm.press_quan = quantile(atm.press, na.rm = TRUE)
    ),
    by = STATE
    ][order(STATE)]
```

**When sorted by state, most states' median is for temp is in the 20s, wind speed is in the 1.5-2.5 range, atm pressure is around 1015.**
